SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization

### 2013-02-21 paper/paper.tex
paper updates.

\begin{abstract}
QuickCheck, developed by Klassen and Hughes, is a powerful library for automatic
test-case generation.  Because QuickCheck performs random testing, it can
discover very large counterexamples.  QuickCheck provides an interface for the
user to write \emph{shrink} functions to attempt to reduce the size of
counterexamples discovered by QuickCheck.  Implementations of shrink often
require significant boilerplate and usually just generate smaller tests than the
discovered counterexample.  Therefore, shrinking can be inefficient and poor at
discovering new, smaller counterexamples.  In this paper, we introduce
\emph{SmartCheck}, an automatic efficient counterexample reduction algorithm.
SmartCheck reduces algebraic data using generic search heuristics to efficiently
find smaller counterexamples.  In addition to shrinking, SmartCheck generalizes
counterexamples by returning a formula that abstracts over individual
counterexamples and make it easier for the programmer to uncover multiple
counterexamples at once.  SmartCheck has been implemented for Haskell and is
freely available.
\end{abstract}

@@

\todo{complexity of each algorithm}
\todo{termination for each algorithm}
\todo{show quantification in intro}
\todo{benchmark feat -- send to author}

@@

In industrial uses, shrinking is essential.  Hughes, the inventor of QuickCheck
and the CEO of Quiviq, a company that commercializes the technology, has noted
that ``without it [shrinking], randomly generated failing cases would often be
so large as to be almost useless.''~\cite{qcjh}.  Hughes~\emph{et al.} give an
extended example in which shrinking is essential in debugging telecom
software~\cite{telecom}.

Typically, effective shrinking on complex data requires a good understanding of
how shrinking in QuickCheck works and the semantics of the property and program
being evaluated.  Bad definitions can be so inefficient that they are unusable.
Shrinking presents a programmer's dilemma: shrinking is used to better
understand a counterexample to understand why a property failed, but to
implement an efficient shrink function in general requires knowing how to
generate counterexamples to the property.

@@

To illustrate, consider the example in Figure~\ref{fig:initial}.\footnote{All
  examples and algorithms in this paper are presented in Haskell.  We strive to
  use idiomatic Haskell, and only rely on the language extensions of existential
  type quantification and scoped type variables.~\cite{haskell98}.}  Data type
\ttp{T} is a product type containing four lists of \ttp{Int16}s and a single
\ttp{Word8}.  For our purposes, it is not important what \ttp{T} models, but for
concreteness, suppose it models the values in four linked lists in an operating
system, together with an 8-bit status register.

Now suppose we are modeling some program that serializes values of type \ttp{T}.
The input to the program satisfies the invariant \ttp{pre}, that the sum of
values in each list of \ttp{Int16}s is less than or equal to 256 (and by
definition, the \ttp{Word8} value is less than or equal to 256).  Assuming this,
we want to show \ttp{post} holds, that the sum of all the values from \ttp{T} is
less than $5 * 256$, where five is the number of fields in \ttp{T}.  At first
glance, the property seems reasonable.  But we have forgotten about underflow;
for example, the value
%
\begin{code}
T (N 10) [N (-20000)] [N (-20000)] [] []
\end{code}
%
\noindent
fails \ttp{prop} (\ttp{==>} is implication from the QuickCheck library).

Despite the simplicity of the example, a typical counterexample returned by
QuickCheck is not quite so simple.  With standard settings, the median
counterexample discovered contains around 70 \ttp{N} constructors, and
counterexamples can contain over 100 constructors.  Thus, it pays to define
\ttp{shrink}!

@@

A more clever programmer that understands the semantics of QuickCheck's
\ttp{shrink} implementation,\footnote{The following approach was suggested to
  the author by John Hughes.} defines a \ttp{shrink} instance as follows:
%
\begin{code}
shrink (T w i0 i1 i2 i3) =
  let xs = shrink (w, i0, i1, i2, i3) in
  let go (w', i0', i1', i2', i3') =
    T w' i0' i1' i2' i3' in
  map go xs
\end{code}
%
\noindent
This ``tuple shrink'' definition does not suffer the same shortcomings: it
shrinks a value even if it contains an empty list, and the combinatorial blowup
of shrink candidates is avoided, since a pair \ttp{(a,b)} is shrunk by
attempting to shrink \ttp{a} while holding \ttp{b} constant, then attempting to
shrink \ttp{b} while holding \ttp{a} constant, and the approach is generalized
to larger tuples.  Still, with this definition, in some cases, shrinking takes
over 20 seconds and returns a result containing over 60
constructors.\footnote{Results are with a GHC-compiled program, using \ttp{-O2},
  on a 4-core 2.7GHz Intel~i7 processor.}

@@

With SmallCheck, a property is checked up to a depth set by the user that
measures how large values are (typically, the depth is a number of constructors
in the value).  For a product type, SmallCheck must check values produced by
taking the cross-product of the type's fields at a given depth.  SmallCheck is
unable to enumerate all possible values of type \ttp{T} to discover a
counterexample, due to combinatorial explosion.  (A related library named Feat
combines some aspects of SmallCheck and QuickCheck~\cite{feat}; we discuss it
more in the related work (Section~\ref{sec:related}).

@@

Finally, even if we could overcome the problems described above with QuickCheck
and SmallCheck, we are left with two others.
\begin{itemize}
  \item A counterexample is usually representative of other counterexamples, but
    that generalization is left to the user to discover for herself.  For
    example, consider the counterexample presented earlier:
%
\begin{code}
T (N 10) [N (-20000)] [N (-20000)] [] []
\end{code}
%
\noindent
For any \ttp{Word8}, the counterexample still holds.  Thus, we can be more
informative and return a quantified formula rather than a single counterexample:
%
\begin{code}
forall x . T x [N (-20000)] [N (-20000)] [] []
\end{code}
%
\noindent
Such a formula has the added benefit of replacing a potentially large sub-term
of a composite data type value with a quantifier, reducing the sheer size of the term.

  \item Second, during testing, it is sometimes helpful to cover the input space
    and by discovering different counterexamples to \ttp{prop}.  In the example
    in Figure~\ref{fig:initial}, one counterexample is fairly representative of
    others, but this is not true in general (e.g., consider properties about
    abstract syntax trees).  Ideally, the testing framework would automatically
    add a predicate characterizing a counterexample as the precondition to
    \ttp{prop} that strengthens the property.  That is, for a specific
    counterexample \ttp{ex}, we have a predicate \ttp{predEx} that holds if its
    argument is another counterexample that is similar to (i.e., some
    generalization of) \ttp{ex}.  Then we test against a new property
%
\begin{code}
prop' a = not (cex a) ==> prop a
\end{code}
%
%% \noindent
%% where \ttp{==>} is QuickCheck's implication operator that discards tests that
%% violate the precondition.
\end{itemize}

\begin{table}[ht]
\footnotesize
  \begin{center}
    \begin{tabular}{|r||c|c|c|c|}
\hline
         & QC (none)  & QC (trunc) & QC (tuple) & SmartCheck\\
\hline \hline
Maximum  & 0.40s, 124 & 139.15s, 90  & 97.53s, 53 & 0.s, 13\\
\hline
Mean     & 0.08s, 69  & 0.73s, 35  & 2.26s, 13  & 0.07s, 6\\
\hline
Median   & 0.07s, 68  & 0.18s, 32  & 1.90s, 13  & 0.07s, 5\\
\hline
    \end{tabular}
  \end{center}
  \caption{Results for 1000 tests of the program in Figure~\ref{fig:initial}
    using interpreted Haskell.}
  \label{table:results}
\end{table}
\todo{table generated with InitEx2 on Feb. 3 --- do a final test with 1000}

\paragraph{SmartCheck}
Motivated by these limitations of QuickCheck and SmallCheck, we developed
\emph{SmartCheck}.  SmartCheck extends QuickCheck to efficiently and generically
find small counterexamples as well as generalize them as described in the
preceding paragraph.  To motivate the benefit of SmartCheck, consider
Table~\ref{table:results}, giving the results for using QuickCheck and
SmartCheck on the program shown in Figure~\ref{fig:initial} (we omit SmallCheck,
since as noted, it is not even feasible for the example).  We show the maximum,
mean, and median amount of time required to generate a counterexample and to
shrink it (and print it to standard out) as well as the corresponding number of
constructors in the final result.

In the first three columns, the results of using QuickCheck to generate
counterexamples are shown.  The first column contains the results of using
QuickCheck without shrinking to provide a baseline.  The next column shows the
result of using QuickCheck with truncating shrink, which each list is truncated
to 10 elements.  Next, we show the results of the ``tuple shrink'' instance.  In
the last column are the results for SmartCheck, which includes the time required
by QuickCheck to generate the original counterexample.  Not only does SmartCheck
execute faster than the other strategies, it has smaller outliers.  In addition,
the size of the counterexamples are significantly smaller.  And we stress that
SmartCheck achieves these results \emph{generically}.

\todo{change max to std dev, look for max info above.}

@@

To present the algorithms in the following sections, we provide some definitions
in the form of methods of a type class.  SmartCheck requires instances for data
being analyzed.  In Section~\ref{sec:implementation}, we describe how instances
for the type class are derived automatically.  Consider the following types and
type class.
%
\begin{code}
type Idx    = Int
type Size   = Int
data SubVal = forall a. SubTypes a => SubVal a

class Arbitrary a => SubTypes a where
  size     :: a -> Size
  index    :: a -> Idx -> Maybe SubVal
  replace  :: a -> Idx -> SubVal -> a
  constr   :: a -> String
  constrs  :: a -> [String]
  baseType :: a -> Bool
  subVals  :: a -> Tree SubVal
\end{code}
%
\noindent
The \ttp{SubTypes} type class requires QuickCheck's \ttp{Arbitrary} as a
super-class.  \ttp{SubTypes} has the following methods:
\begin{itemize}
\item \ttp{size} returns the size of a value,
\item \ttp{index} returns a \emph{sub-value} at an breadth-first index in a
  value,
\item \ttp{replace} replaces a sub-value at a particular focus (returning the
  original value if the index is out-of-bounds),
\item \ttp{constr} returns a string representation of the constructor tagging
  the value,
\item \ttp{constrs} returns the list of all possible constructors from the
  value's type,
\item \ttp{baseType} returns whether the type of the value is an ``interesting
  type'', which generally means it has structure (e.g., a sum or product type).
  We discuss the purpose of base types in Section~\ref{sec:base}.
\item \ttp{subVals} returns a tree of all non base-type sub-values.  A tree is a
  value with the type
%
  \begin{code}
data Tree a = Node \{ rootLabel :: a
                   , subForest :: [Tree a] \}
  \end{code}
\end{itemize}

\noindent
For example, consider a binary tree type:
%
\begin{code}
data T = L | B T T
\end{code}
%
\noindent
and the value \ttp{tree}, labeled with indexes in a breadth-first order:
%
\begin{code}
tree = B\sub{0} (B\sub{1} L\sub{3}
             (B\sub{4} L\sub{6} L\sub{8}))
          (B\sub{2} L\sub{5} L\sub{7})
\end{code}
%
\noindent
Here are example applications of \ttp{SubTypes} methods; in the following, we
show the indexes with respect to the \ttp{tree} value:
%
\begin{code}
size tree = 9

@@

baseType (3 :: Int) = True
baseType L          = False

@@

Figure~\ref{fig:reduction} shows the reduction algorithm.  The function
\ttp{reduce} takes arguments to customize the algorithm's behavior, a
counterexample \ttp{cex} and the property \ttp{prop}.  At the beginning of
execution, we assume that a counterexample has been obtained somehow; our
implementation relies on QuickCheck as a ``front-end'' to generate a
counterexample that the algorithm automatically shrinks.  The reduction begins
at the first proper sub-value of \ttp{cex}; call it \ttp{v}.  When the index
\ttp{idx} becomes out-of-bounds and returns \ttp{Nothing}, the algorithm
terminates.  Otherwise, a list of new random values are generated.
%
\begin{code}
sizedArbitrary :: SubTypes a => a -> IO a
\end{code}
%
\noindent
generates a new value \ttp{v'} having the same type as \ttp{v} and that is
strictly smaller (with respect to the \ttp{size} method) than \ttp{v}.  Just
like QuickCheck's \ttp{arbitrary} method, \ttp{sizedArbitrary} generates
successfully larger counterexamples when generating new values with which to
replace a sub-value.

\ttp{sizedArbitrary} returns a value inside the \ttp{IO} monad since it depends
on a random seed for generation (we could pass in the generator seed, but for
simplicity, let us assume one is generated locally).  The constant
\ttp{scMaxReduce} is the maximum number of tries to discover a new
counterexample by replacing \ttp{v} in \ttp{cex} and testing it.  The result of
\ttp{pass prop cex'} for
%
\begin{code}
pass :: (a -> Property) -> a -> Bool
\end{code}
%
\noindent
holds if \ttp{cex'} satisfies the property \ttp{prop}.  The property may be a
conditional, in which case the value must pass the precondition as well as the
consequent for \ttp{pass} to return \ttp{True}.  If no failure is found, we move
to the next sub-value of \ttp{cex} and continue.  However, if a new smaller
counterexample \ttp{cex'} is found, we start a new breadth-first traversal of
\ttp{cex'}, attempting to shrink it further.  The algorithm is guaranteed to
terminate: intuitively, the measure for the function is that either the index
increases or the size of the counterexample being evaluated decreases.  The
complexity of the algorithm is $\mathcal{O}(n^2)$, where $n$ is the number of
constructors in the counterexample.  The analysis assumes generating new
sub-values and testing them is done in constant time.

@@

\todo{correspond to dimension 0 from\cite{agata}}
The number of constructors in an algebraic data value primarily determines how
large and understandable it is.  In our experience, the size of fields that come
from primitive types, like \ttp{Int} or \ttp{Char}, have less of a role in
understanding a counterexample.  However, a testing algorithm might spend
significant time testing new values in which primitive types are reduces.

@@

Generalizing the scenario, during the reduction algorithm's breadth-first
search through a counterexample \ttp{cex}'s sub-values, we may happen upon a
sub-value \ttp{cex'} that has the same type as \ttp{cex} and fails the property
(while passing any preconditions).  In this case, we can return \ttp{cex'}
directly, and rerun the reduction algorithm on \ttp{cex'}.  In
Figure~\ref{fig:reduce0}, we show an updated reduction algorithm, \ttp{reduceOpt}
that implements this optimization.  The function \ttp{testHole} tests the
current sub-value and if it fails the property, the function tests the sub-value
directly.

@@

SmartCheck's implementation contains flags to allow the user to customize its
behavior.  Three flags that are relevant to the reduction algorithm are the
following:
%
\begin{code}
scMaxReduce :: Int
scMaxSize   :: Int
scMaxDepth  :: Maybe Int
\end{code}
%
\noindent
The \ttp{scMaxReduce} flag controls the number of values generated by the
reduction algorithm for each sub-value analyzed.  \ttp{scMaxSize} controls the
maximum size of values generated to replace sub-values by the reduction
algorithm.  Thus, new sub-values must be strictly smaller than the minimum of
\ttp{scMaxSize} and the size of the sub-value being replaced.  (The size of the
value is with respect to the \ttp{size} method in the \ttp{SubTypes} class.)
Finally, \ttp{scMaxDepth} determines the maximum depth in the counterexample the
reduction algorithm should analyze.  A value of \ttp{Nothing} means that the
counterexample should be exhaustively reduced, as the algorithm is presented in
Figure~\ref{fig:reduction}.  For example, the depth of a list is determined by
its length, and the depth of the binary tree defined in Section~\ref{sec:reduct}
is determined by the function \ttp{depth}:
%
\begin{code}
depth L         = 0
depth (B t0 t1) =
  1 + max (depth t0) (depth t1)
\end{code}
%
\noindent
Of the flags, \ttp{scMaxDepth} is the most important for controlling
efficiency, particularly for large product types with significant ``fan out''.
The number of sub-values of a value product type can grow exponentially with
respect to the depth.  Furthermore, note that as the reduction algorithm
descends further, there is less chance to reduce the size of the value overall,
since smaller and smaller sub-values are replaced.

@@

Small counterexamples make debugging easier, but they are just half the battle.
To go from a specific counterexample to the required fix in a program, the
programmer must have a flash of insight in which she generalizes the
counterexample to a set of counterexamples for which the program and property
fails.  For example, in the introduction, we realized that the specific
counterexample
%
\begin{code}
B [A (-32769)] [] [] []
\end{code}
%
\noindent
is representative of the set of counterexamples in which underflow causes the
sum of the fields to be greater or equal to 64.

The generalization step is an important under-appreciated step in the debugging
process.  For counterexamples that are large, deeply-nested values, the
generalization step can be difficult.  One way we might help the programmer is
to provide her not with a single counterexample but with a quantified formula
that characterizes a set of counterexamples.  In particular, a formula reduces
the noise in favor of the signal by abstracting away portions of large
counterexample that are irrelevant to why it violates the property.

@@

\subsection{Universal Sub-Value Generalization}\label{sec:universal}
An example best explains the first kind of quantified formulas we generate.
Recall the calculator language from Section~\ref{sec:subval}.  Recall the
culprit buried within a counterexample, causing it to result in a
division-by-zero error:
%
\begin{code}
Div (Add (C 7) (C 3)) (Add (C (-5)) (C 5))
\end{code}
%
\noindent
Notice that any value substituted in the dividend results in an error.  We can
quantify the error then as follows:
%
\begin{code}
forall x . Div x (Add (C (-5)) (C 5))
\end{code}
%
\noindent
Not only does the generalization assist the programmer's insight, but it reduces
the sheer size of the counterexample, and it can do so significantly.  We call
the kind of formula just shown \emph{universal sub-value generalization} and is
implemented in SmartCheck.

@@

extrapolate :: SubTypes a
  => ScArgs -> a -> (a -> Property) -> IO [Idx]
extrapolate args cex prop = extrapolate' 1 []
  where
  extrapolate' idx idxs
    | subTrees cex idx idxs
    = extrapolate' (idx+1) idxs
    | Just v <- index cex idx = mkNewVals v
    | otherwise = return idxs
    where
    mkNewVals v = do
      vs <- newVals (scMaxSize args)
                    (scMaxExtrap args) v
      extrapolate' (idx+1)
        (if allFail args cex idx vs prop
           then idx:idxs else idxs)

>>>>

\begin{abstract}
QuickCheck, developed by Klassen and Hughes, is a powerful library for automatic
test-case generation.  Because QuickCheck performs random testing, it can
discover very large counterexamples.  QuickCheck provides an interface for the
user to write \emph{shrink} functions to attempt to reduce the size of
counterexamples discovered.  Hand-written implementations of shrink can be
difficult to write: they can complex and consist of significant boilerplate
code.  Without a good understanding of QuickCheck and the property being tested,
they can be very inefficient, generating poor new candidates, and they may fail
to shrink a counterexample significantly.  In this paper, we introduce
\emph{SmartCheck}, an automatic efficient counterexample reduction algorithm.
SmartCheck reduces algebraic data using generic search heuristics to efficiently
find smaller counterexamples independent of the program or property being
tested.  In addition to shrinking, SmartCheck generalizes counterexamples by
returning a formula that abstracts over individual counterexamples and make it
easier for the programmer to uncover multiple counterexamples at once.
SmartCheck has been implemented for Haskell and is freely available.
\end{abstract}

@@

\todo{benchmark feat -- send to author}

@@

In industrial uses, shrinking is essential.  Hughes, an inventor of QuickCheck
and the CEO of Quiviq, a company that commercializes the technology, has noted
that ``without it [shrinking], randomly generated failing cases would often be
so large as to be almost useless.''~\cite{qcjh}.  Hughes~\emph{et al.} give an
extended example in which shrinking is essential in debugging telecom
software~\cite{telecom}.

Typically, effectively shrinking complex data requires a good understanding of
how shrinking in QuickCheck works and the semantics of the property and program
being evaluated.  Bad definitions can be so inefficient that they are unusable.
Shrinking presents a programmer's paradox: shrinking is used to generate a
smaller counterexample to better understand why a property failed, but to
implement an efficient shrink function often requires a deep enough
understanding of the property to efficiently generate smaller counterexamples to
it!

@@

To illustrate, consider the example in Figure~\ref{fig:initial}.\footnote{All
  examples and algorithms in this paper are presented in
  Haskell~\cite{haskell98}.  We strive to use idiomatic Haskell, and only rely
  on the language extensions of existential type quantification and scoped type
  variables.}  Data type \ttp{T} is a product type containing four lists of
\ttp{Int16}s and a single \ttp{Word8}.  For our purposes, it is not important
what \ttp{T} models, but for concreteness, suppose it models the values in four
linked lists in an operating system, together with an 8-bit status register.

Now suppose we are modeling some program that serializes values of type \ttp{T}.
The input to the program satisfies the invariant \ttp{pre}, that the sum of
values in each list of \ttp{Int16}s is less than or equal to 256 (and by
definition, the \ttp{Word8} value is less than or equal to 256).  Assuming this,
we want to show \ttp{post} holds, that the sum of all the values from \ttp{T} is
less than $5 * 256$, where five is the number of fields in \ttp{T}.  At first
glance, the property seems reasonable.  But we have forgotten about underflow;
for example, the value
%
\begin{code}
T (N 10) [N (-20000)] [N (-20000)] [] []
\end{code}
%
\noindent
satisfies \ttp{pre} but fails \ttp{post} (\ttp{==>} is implication from the
QuickCheck library).

Despite the simplicity of the example, a typical counterexample returned by
QuickCheck can be large.  With standard settings and no shrinking, the median
counterexample discovered contains around 70 \ttp{N} constructors, and
counterexamples can contain over 100 \ttp{N} constructors.  Thus, it pays to define
\ttp{shrink}!

@@

A more clever programmer that understands the semantics of QuickCheck's
\ttp{shrink} implementation,\footnote{The following approach was suggested to
  the author by John Hughes.} defines a \ttp{shrink} instance as follows:
%
\begin{code}
shrink (T w i0 i1 i2 i3) = map go xs
  where xs = shrink (w, i0, i1, i2, i3)
        go (w', i0', i1', i2', i3') =
          T w' i0' i1' i2' i3'
\end{code}
%
\noindent
This ``tuple shrink'' definition does not suffer the same shortcomings: it
shrinks a value even if it contains an empty list, and the combinatorial blowup
of shrink candidates is avoided, since a pair \ttp{(a,b)} is shrunk by
attempting to shrink \ttp{a} while holding \ttp{b} constant, then attempting to
shrink \ttp{b} while holding \ttp{a} constant (and is generalized to larger
tuples).  Still, using this definition, from some initial counterexamples,
shrinking may take over 15 seconds and result in a final counterexample
containing nearly 60 \ttp{N} constructors.\footnote{All results reported in the paper
  are with a GHC-compiled program, using \ttp{-O2}, on a 4-core 2.7GHz Intel~i7
  processor.}  We would like to do better, particularly since the data type and
property are so simple.

@@

For a product type, SmallCheck must check values produced by taking the
cross-product of the type's fields at a given depth.  Unfortunately, SmallCheck
fails long before it reaches a depth required to discover a counterexample, due
to combinatorial explosion.  (A related library named Feat combines some aspects
of SmallCheck and QuickCheck~\cite{feat}; we discuss it more in the related work
(Section~\ref{sec:related}).

@@

Finally, even if we could overcome the problems described above with QuickCheck
and SmallCheck, we are left with two others.
\begin{itemize}
  \item A counterexample is usually representative of other counterexamples, but
    that generalization is left to the user to discover for herself.  For
    example, consider the counterexample presented earlier:
%
\begin{code}
T (N 10) [N (-20000)] [N (-20000)] [] []
\end{code}
%
\noindent
For any \ttp{Word8}, the counterexample still holds.  Thus, we can be more
informative and return a quantified formula rather than a single counterexample:
%
\begin{code}
forall x . T x [N (-20000)] [N (-20000)] [] []
\end{code}
%
\noindent
denoting that for any \ttp{Word8}, the counterexample holds.  Such a formula has
the added benefit of replacing a potentially large sub-term of a composite data
type value with a quantifier, reducing the sheer size of the term.

  \item Second, during testing, it is sometimes helpful to cover the input space
    by discovering different counterexamples to \ttp{prop}.  In the example in
    Figure~\ref{fig:initial}, one counterexample is fairly representative of
    others, but this is not true in general (e.g., consider properties about
    abstract syntax trees).  Ideally, the testing framework would automatically
    add a predicate characterizing a counterexample as the precondition to
    \ttp{prop} that strengthens the property.  That is, for a specific
    counterexample \ttp{ex}, we have a predicate \ttp{predEx} that holds if its
    argument is another counterexample that is similar to (i.e., some
    generalization of) \ttp{ex}.  Then we test against a new property
%
\begin{code}
prop' a = not (cex a) ==> prop a
\end{code}
%
%% \noindent
%% where \ttp{==>} is QuickCheck's implication operator that discards tests that
%% violate the precondition.
\end{itemize}

\begin{table}[ht]
\footnotesize
  \begin{center}
    \begin{tabular}{|r||c|c|c|c|}
\hline
           & QC (none)  & QC (trunc) & QC (tuple) & SmartCheck\\
\hline \hline
Mean       & 0.08s, 69  & 0.53s, 35  & 2.24s, 12  & 0.10s, 7\\
\hline
Median     & 0.07s, 69  & 0.17s, 33  & 1.81s, 13  & 0.10s, 6\\
%% \hline
%% MAD        & 0.02s, 13  & 0.09s, 8   & 0.62s, 3   & 0.02s, 2\\
\hline
95\%       & 0.14s, 100 & 1.76s, 65  & 4.06s, 19  & 0.18s, 12\\
\hline
    \end{tabular}
  \end{center}
  \caption{Results for 1000 tests of the program in Figure~\ref{fig:initial}.}
  \label{table:results}
\end{table}

\todo{get 99\% }

\paragraph{SmartCheck}
Motivated by these limitations of QuickCheck and SmallCheck, we developed
\emph{SmartCheck}.  SmartCheck extends QuickCheck to efficiently and generically
find small counterexamples as well as generalize them as described in the
preceding paragraph.  To motivate the benefit of SmartCheck, consider
Table~\ref{table:results}, giving the results for using QuickCheck and
SmartCheck on the program shown in Figure~\ref{fig:initial} (we omit SmallCheck,
since as noted, it is not even feasible for the example).  Each entry is a pair
containing the number of seconds required for counterexample discovery and
shrinking and the size of the final value obtained (by counting \ttp{N}
constructors).  We show the mean, median, median, and the values at the 95th
percentile values.  (The curves produced by these values are not normal, so we
omit standard the deviation.)

In the first three columns, the results of using QuickCheck to generate
counterexamples are shown.  The first column contains the results of using
QuickCheck without shrinking to provide a baseline.  The next column shows the
result of using QuickCheck with truncating shrink, which each list is truncated
to 10 elements.  Compared with no shrinking, this approximately halves the size
of the median counterexample, but doubles the time taken.  There can be large
outliers, however, in the time taken during shrinking.  Next, we show the
results of the ``tuple shrink'' instance.  While the median amount of time
required is much higher, the size of the resulting value is approximately halved
again.  The reason for the increase in time is because shrinking is very
productive and produces a nonempty list for most values.

In the last column are the results for SmartCheck, which includes the time
required by QuickCheck to generate the original counterexample.  The median time
required by SmartCheck to shrink is 0.02 seconds (SmartCheck's execution time
minus the execution time for QuickCheck without shrinking), and the median size
of a resulting value is six \ttp{N} constructors.  Moreover, SmartCheck has a
very small variance---note the mean and median are nearly identical, meaning
there are very few outliers.  And we stress that SmartCheck achieves these
results \emph{generically}, requiring no shrink definitions to be provided by
the user.

@@

To present the algorithms in the following sections, we provide some definitions
in the form of methods of a \ttp{SubTypes} type class.  SmartCheck requires
instances for data being analyzed.  In Section~\ref{sec:implementation}, we
describe how instances for the type class are derived automatically.
%
\begin{code}
type Idx    = Int
type Size   = Int
data SubVal = forall a. SubTypes a => SubVal a

class Arbitrary a => SubTypes a where
  size     :: a -> Size
  index    :: a -> Idx -> Maybe SubVal
  replace  :: a -> Idx -> SubVal -> a
  constr   :: a -> String
  constrs  :: a -> [String]
  baseType :: a -> Bool
  subVals  :: a -> Tree SubVal
\end{code}
%
\noindent
The \ttp{SubTypes} type class requires QuickCheck's \ttp{Arbitrary} as a
super-class.  \ttp{SubTypes} has the following methods:
\begin{itemize}
\item \ttp{size} returns the size of a value,
\item \ttp{index} returns a \emph{sub-value} at an breadth-first index in a
  value,
\item \ttp{replace} replaces a sub-value at a particular focus (returning the
  original value if the index is out-of-bounds),
\item \ttp{constr} returns a string representation of the constructor tagging
  the value,
\item \ttp{constrs} returns the list of all possible constructors from the
  value's type,
\item \ttp{baseType} returns whether the type of the value is an ``interesting
  type'', which generally means it is user-defined and has positive arity.
  We discuss the purpose of base types in Section~\ref{sec:base}.
\item \ttp{subVals} returns a tree of all non base-type sub-values.  A tree is a
  value with the type
%
  \begin{code}
data Tree a = Node \{ rootLabel :: a
                   , subForest :: [Tree a] \}
  \end{code}
%
\end{itemize}

\noindent
To demonstrate typical evaluations of the methods, consider a binary tree type:
%
\begin{code}
data T = L | B T T
\end{code}
%
\noindent
and the value \ttp{tree}, labeled with indexes in a breadth-first order:
%
\begin{code}
tree = B\sub{0} (B\sub{1} L\sub{3}
             (B\sub{4} L\sub{6} L\sub{8}))
          (B\sub{2} L\sub{5} L\sub{7})
\end{code}
%
\noindent
Here are example applications of \ttp{SubTypes} methods; in the following, we
show the indexes with respect to the value \ttp{tree}:
%
\begin{code}
size tree = 9

@@

baseType (3 :: Int) = True
baseType tree       = False
baseType L          = False

@@

Figure~\ref{fig:reduction} shows the reduction algorithm.  The function
\ttp{reduce} takes flags to customize the algorithm's behavior, a
counterexample \ttp{cex}, and the property \ttp{prop}.  At the beginning of
execution, we assume that a counterexample has been obtained somehow; our
implementation relies on QuickCheck as a ``front-end'' to generate a
counterexample that the algorithm automatically shrinks.  The reduction begins
at the first proper sub-value of \ttp{cex}; call it \ttp{v}.  When the index
\ttp{idx} becomes out-of-bounds and returns \ttp{Nothing}, the algorithm
terminates.  Otherwise, a list of new random values are generated.
%
\begin{code}
sizedArbitrary :: SubTypes a => a -> IO a
\end{code}
%
\noindent
generates a new value \ttp{v'} having the same type as \ttp{v} and that is
strictly smaller (with respect to the \ttp{size} method) than \ttp{v}.  Just
like QuickCheck's \ttp{arbitrary} method, \ttp{sizedArbitrary} generates
successfully larger counterexamples when generating new values with which to
replace a sub-value.

The flag \ttp{scMaxReduce} is the maximum number of tries to discover a new
counterexample by replacing \ttp{v} in \ttp{cex} and testing it.  The result of
\ttp{pass prop cex'} for
%
\begin{code}
pass :: (a -> Property) -> a -> Bool
\end{code}
%
\noindent
holds if \ttp{cex'} satisfies the property \ttp{prop}.  The property may be a
conditional, in which case the value must pass the precondition as well as the
consequent for \ttp{pass} to return \ttp{True}.  If no failure is found, we move
to the next sub-value of \ttp{cex} and continue.  However, if a new smaller
counterexample \ttp{cex'} is found, we start a new breadth-first traversal of
\ttp{cex'}, attempting to shrink it further.

The algorithm is guaranteed to terminate: intuitively, the measure for the
function is that either the index increases or the size of the counterexample
being evaluated decreases.  The algorithm's complexity is
$\mathcal{O}(n^2)$, where $n$ is the number of constructors in the
counterexample, assuming that generating new sub-values and testing them is done
in constant time.

@@

The number of constructors in an algebraic data value primarily determines how
large it is.  In our experience, the size of fields that come from primitive
types, like \ttp{Int} or \ttp{Char}, have less of a role in understanding a
counterexample, so in general, we want to prevent our reduction algorithm from
spending time shrinking values from these types.

@@

Generalizing the scenario, during the reduction algorithm's breadth-first
search through a counterexample \ttp{cex}'s sub-values, we may happen upon a
sub-value \ttp{cex'} that has the same type as \ttp{cex} and fails the property
(while passing any preconditions).  In this case, we can return \ttp{cex'}
directly, and rerun the reduction algorithm on \ttp{cex'}.  In
Figure~\ref{fig:reduce0}, we show an updated reduction algorithm, \ttp{reduceOpt},
that implements this optimization.  The function \ttp{testHole} tests the
current sub-value and if it fails the property, the function tests the sub-value
directly.

@@

SmartCheck's implementation contains flags to allow the user to customize its
behavior.  Three flags that are relevant to the reduction algorithm are the
following:
%
\begin{code}
scMaxReduce :: Int
scMaxSize   :: Int
scMaxDepth  :: Maybe Int
\end{code}
%
\noindent
The \ttp{scMaxReduce} flag controls the number of values generated by the
reduction algorithm for each sub-value analyzed.  \ttp{scMaxSize} controls the
maximum size of values generated to replace sub-values by the reduction
algorithm.  Thus, new sub-values must be strictly smaller than the minimum of
\ttp{scMaxSize} and the size of the sub-value being replaced.  Finally,
\ttp{scMaxDepth} determines the maximum depth in the counterexample the
reduction algorithm should analyze.  A value of \ttp{Nothing} means that the
counterexample should be exhaustively reduced, as the algorithm is presented in
Figure~\ref{fig:reduction}.  For example, the depth of a list is determined by
its length, and the depth of the binary tree defined in Section~\ref{sec:reduct}
is determined by the function \ttp{depth}:
%
\begin{code}
depth L         = 0
depth (B t0 t1) = 1 + max (depth t0) (depth t1)
\end{code}
%
\noindent
Of the flags, \ttp{scMaxDepth} is the most important for controlling
efficiency, particularly for large product types with significant ``fan out''.
The number of sub-values of a value product type can grow exponentially with
respect to the depth.  Furthermore, note that as the reduction algorithm
descends further, there is less chance to reduce the size of the value overall,
since smaller and smaller sub-values are replaced.

@@

Small counterexamples make debugging easier, but they are just half the battle.
To go from a specific counterexample to the required fix in a program, the
programmer must have a flash of insight in which she generalizes the
counterexample to a set of counterexamples for which the program and property
fails.  The generalization step is an important yet under-appreciated step in
the debugging process.  A characterizing formula reduces the noise in favor of
the signal by abstracting away portions of large counterexample that are
irrelevant to why it violates the property.

@@

\subsection{Universal Sub-Value Generalization}\label{sec:universal}
Recall from the introduction the quantified formula
%
\begin{code}
forall x . T x [N (-20000)] [N (-20000)] [] []
\end{code}
%
\noindent
characterizing a set of counterexamples.  Another example using the calculator
language from Section~\ref{sec:subval} is to generalize a counterexample like
%
\begin{code}
Div (Add (C 7) (C 3)) (Add (C (-5)) (C 5))
\end{code}
%
\noindent
by
%
\begin{code}
forall x . Div x (Add (C (-5)) (C 5))
\end{code}
%
\noindent
since any dividend results in divide-by-zero.  Not only do the generalizations
assist the programmer's insight, but they reduce the sheer size of the
counterexample.  We call the kind of formula just shown \emph{universal
  sub-value generalization} and is implemented in SmartCheck.

@@

extrapolate :: SubTypes a
  => ScArgs -> a -> (a -> Property) -> IO [Idx]
extrapolate args cex prop = extrapolate' 1 []
  where
  extrapolate' idx idxs
    | subTrees cex idx idxs
    = extrapolate' (idx+1) idxs
    | Just v <- index cex idx = mkNewVals v
    | otherwise = return idxs
    where
    mkNewVals v = do
      vs <- newVals (scMaxSize args)
                    (scMaxForall args) v
      extrapolate' (idx+1)
        (if allFail args cex idx vs prop
           then idx:idxs else idxs)
